D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-02-14 ootavacuous.tex
Updates to the discussion of hardware dependencies
Change the discussion of hardware dependencies to avoid making the
assumption that an instruction must commit its action and all its
outputs at the same time.

Also remove the paragraph on ordering connected with rfi.  I don't
think it will be needed later.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Dependencies between machine instructions can be understood in terms
of the flow of information within a CPU.
Each instruction has a set of inputs and is the source of a set of outputs,
some of which flow to the inputs for later instructions.
A few examples:
\begin{itemize}
\item	An \co{add} instruction would typically have two inputs
	(the register values to be added together) and two outputs
	(the sum to be stored in a general-purpose register and some
	flag bits---e.g., Zero, Carry, and Overflow---to be stored in
	a status register).
\item	A conditional branch instruction would have only one input:
	the flag bits it tests in the status register.
	A computed branch instruction would have as input the register
	containing the address to branch to.
	For both, the output is the new address to be written to the
	instruction-pointer register.
\item	A memory-load instruction's input is the address to load from,
	and its output is the value obtained by the load, to be stored
	in a register.
\item	A memory-store instruction's inputs are the value to store and
	the address at which to store it; there are no outputs.
\end{itemize}
Of course, some of an instruction's inputs may be immediate constants
taken from the instruction itself rather than outputs from some
earlier instruction.
Note: Hardware dependency analysis considers only the information
flowing \emph{within} a CPU, not information flowing between the
CPU and memory, which is handled separately.

Using this scheme, we say that an instruction $J$ is dependent on
another instruction $I$ when any of $I$'s outputs flow into $J$'s inputs,
perhaps indirectly via some intermediate instructions.
Tracing back the flow of information between instructions, you can
see that any hardware dependency must ultimately emanate from a load
instruction or the thread's initial register values (or possibly some sort of
input instruction, but we will not consider that here).
These are the only sources of truly new information in a thread.

@@

A CPU may start executing an instruction speculatively, but at some
point it must \emph{commit} to a decision either to definitely execute the
instruction's action with some collection of well defined inputs and outputs
or else to abandon it.
Thanks to the law of cause and~effect, a CPU is not able to commit an
instruction until all of its inputs' sources have committed.
(Or at least, all that can affect the action and outputs of the
instruction.
For instance, although no existing architectures do this, there is nothing
in principle to prevent a CPU from committing a \co{multiply}
instruction as soon as either one of its inputs' sources has committed to the
value zero.)
The reason is obvious: An input is subject to change at any time until
its source commits,
the action and outputs aren't fully determined until the inputs
are, and the instruction can't commit until its action and outputs are fully
determined.

This means that, with a few possible exceptions such as the
multiply-by-zero example, an instruction can't commit until all the
instructions it depends on have committed.
And even taking the exceptions into account, if a change in the
outputs of instruction $I$ would lead to a change in the action or outputs of
instruction $J$ then $I$ must commit before $J$.
Thus dependencies force instructions to commit in order, even on
weakly ordered architectures.

However, dependencies aren't the only mechanism capable of forcing
ordering in hardware executions.
The simplest of these is a load reading from a store in another
thread, i.e., the rfe relation.
As mentioned in section~\ref{sec:Simple OOTA Cycle} above, a CPU does
not make the value of a store instruction available for other CPUs to
load until the store commits or some time later.
And after this happens, it takes some time for the information about
the store to travel from that CPU to others, owing to the inescapable
facts that processors have nonzero size and information cannot travel
faster than the speed of light.
Since the load instruction cannot commit until its output (the value it
reads) is fully determined, it must commit after the store it reads
from.

A similarly straightforward mechanism for ordering is conditional or
computed branches.
An instruction executing after such a branch cannot commit
until the CPU has committed to whether the branch will be taken and
if taken, where it will branch to;
until then the CPU cannot know whether the later instruction should
be executed at all.
Therefore instructions following a conditional or computed branch
instruction in an execution must commit after anything the branch
depends on.
(Again, there could theoretically be an exception for a branch that
conditionally jumps to the immediately following instruction.
We can ignore this issue by treating such branches as no-ops.)

Yet another mechanism for ordering crops up when a load reads from a
store in the same thread (the rfi relation).
In this situation the load cannot commit before the sources for the
store's inputs.
Until the source for the store's value commits, the value of the load
isn't fully determined and so the load cannot commit.
And in general, until the source for store's address commits, it's not
fully determined whether the store and load will access the same address.
This argument doesn't hold if the store and load have the \emph{same}
source for their addresses, but another factor is also at play.
Until the address is fully determined, there is a possibility that it
will turn out to be invalid (a NULL pointer, for example), causing the
store instruction to generate an addressing exception.
If this should happen the load might not execute at all, and therefore
it can't commit until the address's source has committed.

@@

Indeed, suppose that abstract execution $A$ is realized by hardware
execution $H$ and $A$ has an OOTA cycle.
This means there are stores $W_i$ in $A$, semantically depending on loads
$\{R_{i,j}\}$ where each of the loads reads from one of the $W_k$
stores in a different thread.
Let $W_i'$ and $R_{i,j}'$ be the hardware instructions these accesses
map to in $H$, if they aren't omitted.
Assuming that at least one of the stores is not omitted,
one of the $W_i'$ instructions, let's say $W_0'$, must commit first.
By the Fundamental Property, one of the loads that $W_0$ depends on,
let's say $R_{0,0}$, is not omitted and $R_{0,0}'$ is ordered before $W_0'$.
But now we have a contradiction:
\begin{itemize}
\item	$W_0'$ commits after $R_{0,0}'$;
\item	$R_{0,0}'$ commits after the store instruction $W_k'$ it reads from;
\item	$W_k'$ commits no earlier than $W_0'$.
\end{itemize}

If all the stores in the OOTA cycle are omitted then all the reads
must be omitted as well.
This amounts to a situation where the entire cycle has been optimized
out of existence by the compiler.
Although we can are unable to prove it, we conjecture that in this
situation there must be another abstract execution which has the
same observable effects as $A$ and is also realized by $H$, but in
which the OOTA cycle does not occur.
Thus there would be no way to tell, merely by observing the effects of
$A$, whether there was an OOTA cycle or not.
For this reason we declare realizations of OOTA cycles in which
all the accesses are omitted to be trivial.

>>>>

Dependencies between machine instructions can be understood in terms
of the flow of information within a CPU.
Each instruction has a set of inputs and is the source of a set of outputs,
some of which flow to the inputs for later instructions.
The inputs determine what an instruction will do.
A few examples:
\begin{itemize}
\item	An \co{add} instruction would typically have two inputs
	(the register values to be added together) and two outputs
	(the sum to be stored in a general-purpose register and some
	condition-code bits---e.g., Zero, Carry, and Overflow---to be
	stored in a status register).
\item	A conditional move instruction would have as inputs the
	condition-code bits to test, the source register value, and the
	target register value; the output would be the value to be stored
	in the target register.
\item	A conditional or computed branch instruction would have as inputs
	the condition-code bits to test, the address of the following
	instruction (to be used if the condition is false) and the
	destination address (to be used if the condition is true).
	The output is the new address to be written to the
	instruction-pointer register.
\item	A memory-load instruction's input is the address to load from,
	and its output is the value obtained by the load, to be stored
	in a register.
\item	A memory-store instruction's inputs are the value to store and
	the address at which to store it; there are no outputs.
\end{itemize}
Note: Hardware dependency analysis considers only the information
flowing \emph{within} a CPU, not information flowing between the
CPU and memory, which is handled separately as part of the action
taken by an instruction.

Using this scheme, we say that an instruction $J$ is dependent on
another instruction $I$ when any of $I$'s outputs flow into $J$'s inputs,
perhaps indirectly via some intermediate instructions.
Tracing back the flow of information between instructions, you can
see that any hardware dependency must ultimately emanate from a load
instruction or the thread's initial register values (or possibly some sort of
input instruction, but we will not consider that complication here).
These are the only sources of truly new information in a thread.

@@

A CPU may start executing an instruction speculatively, but at some
point it must \emph{commit} to a decision either to definitely execute the
instruction's action with some collection of well defined inputs and outputs
or else to abandon it.
Thanks to the law of cause and~effect, a CPU is not able to commit an
instruction or its outputs until the relevant inputs' sources have committed.
The reason is obvious: An input is subject to change at any time until
its source commits to its value,
and the instruction and its outputs can't commit until the CPU has fully
determined what it will do and what they will be.
(On the other hand, the CPU need not wait for inputs that won't affect
the instruction's results.
For instance, a conditional move instruction wouldn't need to wait for
the source register input to be determined once it knew that the
condition was definitely false.
And, although no existing architectures do this, there is nothing
in principle to prevent a CPU from committing a \co{multiply}
instruction as soon as either one of its inputs' sources has committed to the
value zero.)

Of course, an instruction's inputs don't all have to be outputs from
earlier instructions; some of them may be immediate constants present
in the instruction itself.
In this case there is no need to wait for those inputs because
their values are fully determined from the start.

The overall effect of these hardware dependencies is that if a change to an
output of instruction $I$ would lead to a change in the action or outputs of
a later instruction $J$, then the CPU must commit $I$'s output before
committing $J$'s action and outputs.
Thus dependencies force instructions to commit in order, even on
weakly ordered architectures.

However, dependencies aren't the only mechanism capable of forcing
ordering in hardware executions.
The simplest alternative is a load reading from a store in another
thread, i.e., the rfe relation.
As mentioned in section~\ref{sec:Simple OOTA Cycle} above, a CPU does
not make the value of a store instruction available for other CPUs to
load until the store commits or some time later.
And after this happens, it takes some time for the information about
the store to travel from that CPU to others, owing to the inescapable
facts that processors have nonzero size and information cannot travel
faster than the speed of light.
Since the load instruction cannot commit until its output (the value it
reads) is fully determined, it must commit after the store it reads
from.

A similarly straightforward mechanism for ordering is conditional or
computed branches.
An instruction executing after such a branch cannot commit
until the CPU has committed to whether the branch will be taken and
if taken, where it will branch to (that is, the output value it will
send to the instruction-pointer register);
until then the CPU cannot know whether the later instruction should
be executed at all.
Therefore instructions following a conditional or computed branch
instruction in an execution must commit after the branch, and hence
after the source for the branch's condition or destination input.
(There could theoretically be an exception for a branch that
conditionally jumps to the immediately following instruction.
We can ignore this possibility by treating such branches as no-ops.)

@@

Indeed, suppose that abstract execution $A$ is realized by hardware
execution $H$ and $A$ has an OOTA cycle.
This means there are atomic stores $W_i$ in $A$,
semantically depending on atomic loads
$\{R_{i,j}\}$ where each of the loads reads from one of the $W_k$
stores in a different thread.
Let $W_i'$ and $R_{i,j}'$ be the hardware instructions these accesses
map to in $H$, if they aren't omitted.
Assuming that the stores are not all omitted,
one of the $W_i'$ instructions, let's say $W_0'$, must commit first.
By the Fundamental Property, one of the loads that $W_0$ depends on,
let's say $R_{0,0}$, is not omitted and $R_{0,0}'$ is ordered before $W_0'$.
But now we have a contradiction:
\begin{itemize}
\item	$W_0'$ commits after $R_{0,0}'$;
\item	$R_{0,0}'$ commits after the store instruction $W_k'$ it reads from;
\item	$W_k'$ commits no earlier than $W_0'$.
\end{itemize}

If all the stores in the OOTA cycle are omitted then all the reads
must be omitted as well.
In effect, the entire cycle has been optimized
out of existence by the compiler.
Although we are unable to prove it, we conjecture that in this
situation there must be another abstract execution which has the
same observable effects as $A$ and is also realized by $H$, but in
which the OOTA cycle does not occur.
Thus there would be no way to tell, merely by observing the effects of
$A$, whether there was an OOTA cycle or not.
For this reason we declare realizations of OOTA cycles in which
all the accesses are omitted to be trivial.
